Nextstrain: real-time tracking of pathogen evolution

### 2018-04-17 whitepaper.tex
remove 11/12 usages of the passive voice

The regularly updated nature and rapidity of these analyses is crucial to the monitoring and understanding of pathogen epidemiology and evolution.
Sequencing times and costs are continually dropping, with on-the-ground sequencing being deployed during the recent West African Ebola and Zika epidemics \citep{quick2016real,faria2017epidemic}.
However, this speed of sequencing must be complemented with rapid methods by which to analyse, interpret, and disseminate results.

Nextstrain consists of three components:
\textit{Fauna} is a collection of Python scripts to maintain a database representative of all available sequences and related metadata, sourced from public repositories such as NCBI (www.ncbi.nlm.nih.gov), GISAID (www.gisaid.org) and ViPR (www.viprbrc.org), as well as GitHub repositories and other sources of genomic data.
\textit{Augur} is a suite of tools to perform phylodynamic analysis \citep{voltz2013}, including subsampling, alignment, phylogenetic inference, temporal dating of ancestral nodes and discrete trait geographic reconstruction, including inference of the most likely transmission events.
This leverages the maximum likelihood analyses implemented in TreeTime \citep{sagulenko2017treetime}, allowing a full analysis of the entire Ebola epidemic ($n$=1581 genomes) in under 2 hours on a modern laptop.
These scripts separate generic core functionality from a light pathogen-specific layer such that they are easily co-opted to different pathogens.
Outputs from the pipeline are uploaded online and immediately available through the visualisation software \textit{Auspice}, located at \href{http://nextstrain.org}{nextstrain.org}.
This approach is similar in concept to Nextflu \citep{neher2015nextflu} however extended and generalised to different viral pathogens.
There is a growing need for surveillance of non-influenza viruses \citep{tang2017global}, and Nextstrain is able to be extended to most outbreaks with readily accessible genomic data, although we note the potential for recombination or slow mutation rate to confound phylogenetic signal.

@@

\subsection*{Joint temporal and spatial visualisation}
Conveying understanding of pathogen evolution through space and time involves filtering large amounts of data into forms that can be easily reasoned with.
Auspice employs linked views into different facets of the data, whereby changes in one view are reflected in all others.
These views allow simultaneous interrogation of phylogenetic and spatial relationships, with additional data such as genotype or serotype expressed through colourings (Figure \ref{nextstrain}).
This is coupled with an interactive time slider to see how the pathogen has evolved and spread over the course of the epidemic.
By animating the temporal dimension, a high level overview of how the entire outbreak unfolded is quickly gained.
This approach both communicates the geographical spread of the epidemic alongside the underlying genomic data that supports this geographic reconstruction.

Maximum likelihood ancestral state reconstruction of discrete traits such as country or region of isolation allows identification of probable transmission events given the sampled data, together with a marginal likelihood for each event.
Inferred transmissions are displayed as colours on the tree and each transmission is also drawn on the map.
It is important to make known the confidence of any such reconstruction, and we do this in two ways: by matching colour saturation to the confidence of that trait, and by displaying all relevant information when one hovers over the corresponding branch or isolate on the tree.

@@

For many pathogens, the emergence and spread of gain-of-function variants is a grave concern.
For instance, China has experienced seasonal epidemics of influenza A/H7N9 over the past five years, however human to human transmission is thought to be rare, with the majority of human cases being spillover from poultry.
The threat of mutations which facilitate human to human transmission is of extreme concern, as H7N9 has a mortality rate of around $30\%$ \citep{li2014epidemiology}.
% \textit{In vitro} experiments have identified three amino acid mutations in the hemagglutinin (HA) gene which confer a switch in specificity for human-type receptors by promoting binding to human trachea epithelial cells \citep{devries2017three}.
Nextstrain allows monitoring of experimentally determined mutations, such that interventions and control measures may be implemented where appropriate.

@@

Analysis of such outbreaks relies on public sharing of data, and Nextstrain has the ability to automatically update as new sequences from a range of public databases and repositories appear.
Scientists are justifiably hesitant to cede control of their data, and we try to address these concerns by preventing access to the raw genome sequences, and by clearly indicating the source of each sequence.
Derived data, such as phylogenetic trees, metadata and screenshots may be downloded, and private metadata may be added by users by simply dragging a CSV file onto the browser.
We believe this strikes a compromise between keeping certain data private and allowing the dissemination of results important to the wider scientific community, thereby encouraging collaboration between scientists.
Genomic epidemiology has the potential to inform the public, health organisations and scientists alike, a potential realised by sharing of data in real-time rather than retrospectively \citep{croucher2015application}.

>>>>

The regularly updated nature and rapidity of these analyses is crucial to the monitoring and understanding of pathogen epidemiology and evolution.
Sequencing times and costs are continually dropping, with on-the-ground sequencing used during the recent West African Ebola and Zika epidemics \citep{quick2016real,faria2017epidemic}.
Rapid methods by which to analyse, interpret, and disseminate results must complement this sepeed of sequencing.

Nextstrain consists of three components:
\textit{Fauna} is a collection of Python scripts to maintain a database representative of all available sequences and related metadata, sourced from public repositories such as NCBI (www.ncbi.nlm.nih.gov), GISAID (www.gisaid.org) and ViPR (www.viprbrc.org), as well as GitHub repositories and other sources of genomic data.
\textit{Augur} is a suite of tools to perform phylodynamic analysis \citep{voltz2013}, including subsampling, alignment, phylogenetic inference, temporal dating of ancestral nodes and discrete trait geographic reconstruction, including inference of the most likely transmission events.
This leverages the maximum likelihood analyses implemented in TreeTime \citep{sagulenko2017treetime}, allowing a full analysis of the entire Ebola epidemic ($n$=1581 genomes) in under 2 hours on a modern laptop.
These scripts separate generic core functionality from a light pathogen-specific layer such that they are easily co-opted to different pathogens.
Visualisation of these analysis is available through \href{http://nextstrain.org}{nextstrain.org}.
This approach is similar in concept to Nextflu \citep{neher2015nextflu} however extended and generalised to different viral pathogens.
There is a growing need for surveillance of non-influenza viruses \citep{tang2017global}, and Nextstrain is able to be extended to most outbreaks with readily accessible genomic data, although we note the potential for recombination or slow mutation rate to confound phylogenetic signal.

@@

\subsection*{Joint temporal and spatial visualisation}
Conveying understanding of pathogen evolution through space and time involves filtering large amounts of data into forms that can be easily reasoned with.
Multiple views into different facets of the data are presented and remain in sync as one interacts with the data.
This allows simultaneous interrogation of phylogenetic and spatial relationships, with additional data such as genotype or serotype expressed through colourings (Figure \ref{nextstrain}).
This is coupled with an interactive time slider to see how the pathogen has evolved and spread over the course of the epidemic.
By animating the temporal dimension, a high level overview of how the entire outbreak unfolded is quickly gained.
This approach both communicates the geographical spread of the epidemic alongside the underlying genomic data that supports this geographic reconstruction.

Maximum likelihood ancestral state reconstruction of discrete traits such as country or region of isolation allows identification of probable transmission events given the sampled data, together with a marginal likelihood for each event.
Internal node colours indicate inferred transmissions on the tree, which are also drawn as links between demes on the map.
It is important to make known the uncertainty of such reconstructions.
The confidence is conveyed by matching colour saturation to the confidence of that trait, and by displaying all relevant information when one hovers over the corresponding branch or isolate on the tree.
Despite this, sampling bias can obscure transmission links, and in cases of extreme bias we have chosen not to display the inferred states.

@@

For many pathogens, the emergence and spread of gain-of-function variants is a grave concern.
For instance, China has experienced seasonal epidemics of influenza A/H7N9 over the past five years.
Despite no known cases of human to human transmission, the high mortality rate of $~30\%$ \citep{li2014epidemiology} makes the threat of mutations which facilitate human to human transmission of extreme concern (e.g. see \citep{devries2017three}).
Nextstrain allows monitoring of experimentally determined mutations, informing appropriate interventions and control measures.

@@

Analysis of such outbreaks relies on public sharing of data, and Nextstrain has the ability to automatically update as new sequences from a range of public databases and repositories appear.
Scientists are justifiably hesitant to cede control of their data, and we try to address these concerns by preventing access to the raw genome sequences, and by clearly indicating the source of each sequence.
Derived data, such as phylogenetic trees, metadata and screenshots are available, and private metadata added via CSV files.
We believe this strikes a compromise between keeping certain data private and allowing the dissemination of results important to the wider scientific community, thereby encouraging collaboration between scientists.
Genomic epidemiology has the potential to inform the public, health organisations and scientists alike, a potential realised by sharing of data in real-time rather than retrospectively \citep{croucher2015application}.
